\name: Combining Bottom-up and top-down approaches through graph learning over interaction networks for drug-target-interaction prediction

### 2021-03-22 paper/main.tex
Updated introduction, added some TODO

\cite{Survey2018}\\
In history, traditional remedies, that were known for their medicinal
properties lead to drugs by extraction of the functional
ingredients. Alternatively, characteristics and features of potential
drugs were detected by accident like in the case of penicillin. More
recently, biological drug targets can be found \textit{in silico}
through discovery of suitable computational predictors.

@@

With the rapidly growing field of machine learning approaches and
their application to bioscientifical problems in the realm of
bioinformatics, different kinds of data, such as long DNA sequences
could be utilized for feature generation, while rapid advances were
made. Almost all state of the art models for drug-target-interaction
prediction were based on the usage of neural networks with increasing
size.\todo{Needs ref, or reformulate}

Only recently, the technique of graph learning was introduced by
\citet{GCNConv} through graph convolution algorithms, and improved and
altered under usage of different kernels \cite{ChebConv, ARMAConv},
attention mechanisms \cite{GATConv}, random walks \cite{APPNPConv},
and mixtures of both \cite{SAGEConv}. While based on diverse systems,
they can be relevant for testing distinct hypothesis for given
graphs. While convolutional filters are suitable for finding patterns
among the the given graph, attention mechanisms are more relevant for
discovery of important regions within. Lately, graph learning
approaches found application for computing compound representations
for DTI prediction.

@@

\todo[inline]{Introduction}
Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we call \textit{hub proteins}) have significantly more known
interactions with drugs than others. In the STITCH database, $5\%$ of
the proteins have $40\%$ of the interactions, and similar
distributions are present in the Yamanishi and Drugbank
\cite{Drugbank2007, Drugbank2017} datasets; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (e.g.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed).

@@

\section{Methods}
\subsection{Problem Description}
\name aims to solve the following problem: for a given drug and a
given protein we want to determine whether those interact or not.  We
do not differentiate between types of interaction such as activation
and inhibition, and do not predict the strength of the interaction.
If we additionally assume that our knowledge is complete and all
drug--protein pairs without a known interaction do not interact, we
can formulate the problem as a binary classification task.

\subsection{Datasets}
We obtain a dataset consisting of 12,884 human proteins with over
340627 links from STRING \citep{STRINGv10}. For the drug-target
interactions, we use 229,870 links from the STITCH database
\citep{STITCHv5}. As both STRING and STITCH provide confidence scores
for each association, we filtered them as advised by a threshold of
$700$, therefore retaining only high-confidence interactions.

@@

We only use proteins in our analysis that have at least one link in
either STITCH or STRING, and drugs with at least one side effect and
one existing target. Therefore, the intersection between these
resources yields 1,428 drugs and 7,368 human proteins with 32,212
interactions for the training phase. We provide links to and methods
for obtaining and processing the necessary data on Github.

For comparative evaluation, we use the gold standard dataset
introduced by \citet{Yaminishi2008} consisting of 1,923 interactions
between 708 drugs and 1,512 proteins.

@@

\subsubsection{Half-twin neural networks and feature transformation}

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with a
contrastive loss using cosine similarity.  A half-twin neural network
aims to learn a similarity between two embeddings of the same
dimension.  The precomputed embeddings may have different
dimensionality and we first process them using a learnable feature
transformation (LFT) network, i.e., a fully connected neural network
layer which takes as input an embedding and outputs a representation
of a particular size. An example structure for both types of features
can be found in Figure~\ref{fig:HalfTwinNetwork}. We use this LFT to
reduce the representation size of the embeddings for drugs and
proteins separately.
% While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we
The use of an LFT enables flexible experimentation as both ontology
and molecular feature for both drugs and proteins are reduced to
the same dimensionality for varying sizes of inputs; this allows for a
high amount of modularity across different experimental setups by
adding different kinds of features into the model. Additionally, the
generated features may be used for other tasks. % ; for
% example, the ontology LFT can be reused for a variety of DL2vec-based
% features with respect to other ontologies and hypotheses
We follow the results of \textit{DL2vec} \cite{} and use as
activation function $\sigma := \mathrm{LeakyReLU}$ which leads to
improved performance compared to other activation functions.

@@

We include these molecular and ontology-based sub-models within a
graph neural network (GNN). The graph underlying the GNN is based on
the protein--protein interaction (PPI) graph. The PPI dataset is
represented by a graph $G=(V,E)$, where each protein is represented by
a vertex $v\in V$, and each edge $e\in E\subseteq V\times V$
represents an interaction between two proteins. Additionally, we
introduce a mapping $x:V\rightarrow\mathbb{R}^{d}$ projecting each
vertex $v$ to its node feature $x_v := x(v)$, where $d$ denotes the
dimensionality of the node features.

% As described before, graph convolution has shown significant
% performance increase in a variety of tasks. While there are various
% methods out there we will only introduce the most basic one here. 
A graph convolutional layer \cite{GCNConv} consists of a learnable
weight matrix followed by an aggregation step, formalized by
\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}
where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the
adjacency matrix with added self-loops for each vertex, $D$ is
described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal
matrix displaying the degree of each node, and $\Theta$ denotes the
learnable weight matrix. Added self-loops enforce that each node
representation is directly dependent on its own preceding one. The
number of graph convolutional layers stacked equals the radius of
relevant nodes for each vertex within the graph.

The update rule for each node is given by a message passing scheme
formalized by
\begin{equation}
	\mathbf{x}^{\prime}_i = \mathbf{\Theta} \sum^{N}_{j}
	\frac{1}{\sqrt{\hat{d}_j \hat{d}_i}} \mathbf{x}_j
\end{equation}
where both $\hat{d}_i, \hat{d}_j$ are dependent on the edge weights
$e_{ij}$ of the graph. With simple, single-valued edge weights such as
$e_{ij}=1 \text{ }\forall (i,j)\in E$, all $\hat{d}_i$ reduce to
$d_i$, i.e., the degree of each vertex $i$. We denote this type of
graph convolutional neural layers with \textsc{GCNConv}.

While in this initial formulation of a GCNConv the node-wise update
step is defined by the sum over all neighbouring node representations,
we can alter this formulation to other message passing schemes.  We
can rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$, and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We denote
this generalized layer type as \textsc{GENConv} following the
notation of PyTorch Geometric \cite{PytorchGeometric}.  While the
reordering is mainly import for numerical stability, this alteration
also addresses the vanishing gradient problem for deeper convolutional
networks \cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale and highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020}.  The residual blocks are reusable and can be stacked
multiple times. The structure of the GNN architecture we use is shown
in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{Combined prediction model}
Combining half-twin and graph convolutional neural networks, we map
all protein representations to their respective node features,
initializing the graph convolutional update steps. The resulting
representations are used for a similarity prediction similar
to\todo{Similar, or identical to? Needs to describe the exact model
  used} the models presented in figure ~\ref{fig:HalfTwinNetwork}.
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph for the similarity
computation. A more in-depth image of the overall architecture,
combining both feature types is depicted in figure
~\ref{fig:FullModelAllLayers}.

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the number of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered subset the ratio
\begin{equation}
	w:= \frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
	\label{weight}
\end{equation}
and consequently needs to be compensated in the computed loss function
and appropriate metrics for the evaluation.

Therefore, we weight all positive drug-protein pair samples with this
ratio by introducing the following loss function with respect to 
binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for a given prediction $x$ and target $y$, and positive weight $w$
defined by equation \eqref{weight}. We average this loss among all
drug-protein pairs in the training set, leading to a stable
environment for the \textit{Adam} optimization algorithm
\citep{Adam2014}. We implemented a 5-fold cross validation split among
the proteins. Furthermore, we used early stopping in the training
process.

To find the best hyperparameter configuration for the proposed model,
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
with respect to our evaluation metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search. Starting from na\"ive, shallow
feature transformations with an embedding size of $10$, we scaled the
network up to residual structures with up to $10$ hidden layers
leading to embeddings of size $4000$, testing different network widths
and learning rates for each configuration.

\subsection{Evaluation and metrics}
\label{sec:evaluation}
To assess each model, we compute a variety of common metrics for
binary classification. As the datasets are highly imbalanced, we use
the area under the receiver operating characteristic curve
(AUROC) on training, validation and testing split. % We
% calculated the true positive rate (TPR), false positive rate (FPR), and
% precision score. We will further compute true positives (TP), false
% positives (FP), false negatives (FN) and finally true negatives (TN).

We calculate the AUROC by computing true positive rate at various
false positive rate thresholds and use trapezoidal approximations to
estimate the area under the curve. We refer to this measure as
$\textrm{MacroAUC}$.

\todo[inline]{For RH: check this again later:}
We also calculate the $\textrm{MicroAUC}$ score. For given lists $D$
and $P$ of drugs and proteins, respectively, and a set of known
interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated
as the average per entity (macro) \textit{AUROC} score. Specifically,
for protein-centric score, this can be formalized as: given labels
$l:D\times P \rightarrow \{0,1\}$ and predictions $y:D\times P
\rightarrow [0,1]$, we define
\begin{equation*}
  \textrm{MicroAUC}_p := \underset{p\in P}{mean}\left(\left\{
      \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\})
    \right\}\right)
\end{equation*}

Drugs and proteins can be interchanged in this formulation, and we
refer to the different measures as protein-centric microAUC
($MicroAUC_p$) and a drug-centric microAUC ($MicroAUC_d$). In some
cases, the $MicroAUC$ score may not be defined because in some
datasets some proteins or drugs have no interactions, leading to an
infeasible $TPR=0$ for all thresholds and an undefined $AUROC$ score
for that entity. For those entities, we impute the $MicroAUC$
interpolating linearly, by using the accuracy for this
subset\todo{Last sentence not clear, can you give the formula instead?}.

@@

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank drugs that likely target this protein. \name combines two types
of features: structural information for drugs and proteins that can be
used to determine if the drug and protein physically interact, and
information about phenotypic effects of drugs and changes in protein
function that may ``localize'' on an interaction network (i.e.,
neighboring nodes will share some of these features or are
phenotypically similar).  As structural features, \name uses
structural representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020} and
ontology-based annotations of drugs and proteins.

We construct a graph with proteins as nodes and protein-protein
interactions as edges, mapping the protein features to each target as
node features. \name then propagates information among the PPI network
utilizing graph convolutional steps, calculates the similarity of drug
and protein representations, and predicts whether there is an
interaction. The full workflow scheme is depicted in
figure~\ref{fig:ModelWorkflow}

@@

\begin{table*}[!tpb]
  \centering
	\begin{tabular}{|p{2.2cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|}
		\hline
		(a) STITCH results&\multicolumn{4}{c|}{PPI graph}\\
		&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
		&Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$\\
		\hline
		MolPred&$0.69$&$0.65$&$0.69$&$0.67$\\
		\hline
		OntoPred&$0.88$&$0.87$&$0.92$&$0.93$\\
		\hline
		\name (MolPred + OntoPred) & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$\\
		\hline
		\hline
		(b) Yamanishi results&\multicolumn{4}{c|}{PPI graph}\\
		&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
		&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		\name (MolPred + OntoPred) & $0.83$ & $0.82$&$0.84$&$0.84$\\
		\hline
	\end{tabular} \medskip

@@

	\caption{(a) + (b) Results for \name on STITCH and Yamanishi dataset evaluated with a 5-fold cross-validation. We hereby denote the molecular feature based predictor with \textit{MolPred}, while abbreviating the ontology based, top-down predictor with \textit{OntoPred}. (c) + (d) Results for various state of the art (c) drug--target interaction prediction methods on Yamanishi dataset and (d) drug--target affinity prediction methods on BIOSNAP dataset, evaluated on their original and the protein cross-validation splitting scheme, approximately reproducing the results of MolTrans \textbf{CITATION}.}
	\label{tab:Results}
\end{table*}

@@

We trained, validated and finally tested all considered models on the
STITCH dataset using a 5-fold cross-validation over a protein split;
we then selected the best-performing models (with respect to
\textit{MacroAUC} and $MicroAUC_p$\todo{which is it, macro or micro?},
see Section \ref{sec:evaluation}) and evaluate them in a 5-fold
protein-split cross-validation on the Yamanishi benchmark dataset to
avoid validation overfitting and yield more realistic testing
results. To evaluate the influence of the different features
separately, and to determine whether they ``localize'' on the PPI
graph (and therefore can be exploited successfully by the graph neural
networks), we train and evaluate models with different types of
features, and with and without inclusion of the PPI graph,
separately. We comparing the molecular (MolPred) and phenotype-based
(OntoPred) prediction model, and a combination of both where we
concatenate both types of features.  Table
\ref{tab:Results}\todo{check -- not currently in the table} shows the
results of these experiments.

>>>>

In history, traditional remedies, that were known for their medicinal
properties lead to drugs by extraction of the functional
ingredients. Alternatively, characteristics and features of potential
drugs were detected by accident like in the case of penicillin. More
recently, biological drug targets can be found \textit{in silico}
through discovery of suitable computational predictors.

@@

With the rapidly growing field of machine learning approaches and
their application to bioscientifical problems in the realm of
bioinformatics, different kinds of data, such as long DNA sequences
could be utilized for feature generation, while rapid advances were
made.

Only recently, the technique of graph learning was introduced by
\citet{GCNConv} through graph convolution algorithms, and improved and
altered under usage of different kernels \cite{ChebConv, ARMAConv},
attention mechanisms \cite{GATConv}, random walks \cite{APPNPConv},
and mixtures of both \cite{SAGEConv}. While based on diverse systems,
they can be relevant for testing distinct hypothesis for given
graphs. While convolutional filters are suitable for finding patterns
among the given graph, attention mechanisms are more relevant for
discovery of important regions within. Lately, graph learning
approaches found application for computing compound representations
for DTI prediction.

These approaches can be classified into top-down and bottom-up methods. Top-down or network approaches hereby start form the observable, phenotypical characteristics, such as side-effects, associated diseases and indications represented by knowledge graphs or ontologies, induced by a drug and infer targets based on the likely molecular mechanisms that result in these phenotypes. On the other hand, bottom-up approaches start from molecular features, such as molecular structure, molecular fingerprints, secondary structure and contact predictions and others, deriving predictions from suitable correlations between these features.

However, both bottom--up and top--down approaches to drug--target interaction prediction contain some limitations that are not solvable within themselves. Generally, it is very challenging for bottom-up methods to accurately predict, whether there a structure is binding to a protein, as not only molecular structure, but also binding sites and molecular forces are crucial for accurate prediction. Thus, finding similarities for dissimilar drugs affecting the same targets is problematic but crucial. Furthermore, phenotypic features contain relevant information for DTI prediction \cite{Campillos2008} that are complementary to knowledge gained from molecular properties. While bottom-up representations possibly contain all information needed for accurate DTI prediction, we rely on indirect information for better generalization. Contrary, top-down approaches lack the ability to spot and cope with small differences within the molecular drug structure, thus both incapable to predict on novel entities and derive similarity from structural correlation.

In order to design representations concerning both top-down and bottom-up features for proteins and drugs,respectively, we utilize protein--protein interaction (PPI) networks which have shown great results in protein function prediction \cite{Vazquez2003} and antiviral drug--target discovery \cite{Ackerman2019}. 
However, only few applied this context to the task of DTI prediction in humans. The method of graph convolutional neural networks has been applied successfully to PPI networks \cite{Zitnik2017}, granting valuable information for medical diagnostics, closely related to phenotypic information.

Eventually, we further address potential biases and skews within DTI prediction datasets. \todo[inline]{Introduction}
Within DTI prediction, there are potential biases resulting from the
underlying datasets \citep{Pahikkala2014}. First, novel drugs are
often designed by altering non-functional components of a drug,
leading to two and more very similar drugs designed to target the same
proteins \cite{Overington2006}. This can result in a bias when it
leads to hidden duplicates that can distribute among the train/test
split, resulting in a better (measured) predictive performance than
would be expected when the model is applied to identify drugs that
target a protein for which no drugs yet exist. Second, some proteins
(which we call \textit{hub proteins}) have significantly more known
interactions with drugs than others. In the STITCH database, $5\%$ of
the proteins have $40\%$ of the interactions, and similar
distributions are present in the Yamanishi and Drugbank
\cite{Drugbank2007, Drugbank2017} datasets; preferentially predicting
these proteins may increase predictive performance while not
reflecting the actual performance when applied to a new protein (e.g.,
a protein for which no interactions are known). These differences in
the number of drugs targeting certain proteins may be the result of
study bias where more ``valuable'' proteins have more drugs designed
to target them due to their involvement in more common diseases (or
diseases for which drugs can be more profitably marketed).

\cite{Survey2018}\\

Almost all state of the art models for drug-target-interaction
prediction were based on the usage of neural networks with increasing
size.\todo{Needs ref, or reformulate}

@@

@@

\section{Methods}
\subsection{Problem Description}
The issue of predicting drug-target interactions can be described
quite briefly: For a given drug and a given protein we want to
determine whether those interact or not.  We do not differentiate
between types of interaction such as activation and inhibition, and do
not predict the strength of the interaction.  If we additionally make
the closed world assumption, i.e., assume that our knowledge is
complete and all drug--protein pairs without a known interaction do
not interact, we can formulate the problem as a binary classification
task.

\subsection{Datasets}
We obtain a dataset consisting of 12884 human proteins with over 340627 links from STRING
\citep{STRINGv10}. For the drug-target interactions, we
fetched 229870 links from the STITCH database
\citep{STITCHv5}. As both STRING and STITCH provide confidence scores
for each association, we filtered them as advised by a threshold of
$700$, therefore retaining only high-confidence interactions.

@@

We only use proteins in our analysis that have at least one link in
either STITCH or STRING, and drugs with at least one side effect and
one existing target. Therefore, the intersection between these
resources yields 1,428 drugs and 7,368 human proteins with 32212 interaction links for the training phase. We provide links to and methods for obtaining and processing
the necessary data on Github.

For comparative evaluation, we use the gold standard dataset
introduced by \citet{Yamanishi2008} consisting of 1923 interactions between 708 drugs and 1512 proteins.

@@

\subsubsection{Half-twin neural networks and modular learnable feature transformation}

\todo[inline]{Looking at the code, I wonder if what is implemented is
  actually a 
   ANN; it does not look as if the weights are
  shared between the two networks, so looks more like a twin or
  half-twin network? Siamese would be sim(f(e1),f(e2)) but the
  implementation seems to be sim(f(e1), g(e2)).}

As we want to learn from the similarity of drug side effects and
protein phenotypes, we use a deep half-twin neural network with
a contrastive loss using cosine similarity.
A half-twin neural network aims to learn a similarity between two
embeddings.

% learning a high-dimensional embedding emphasizing this identity by
% forcing a similarity between these embeddings. On the other hand we
% built a deep neural network for the molecular structure based
% features, also benefiting from the siamese network
% architecture. Computing the similarity between two representations,
% allows for a variety of different methods.  
% However, we decided for
% the cosine similarity measure being invariant to scaling.

Therefore, the precomputed embeddings are run through a regular,
neural learnable feature transformation (LFT) network, which also
reduces the eventual representation size for drugs and proteins
separately. An example structure for both types of features can be
found in Figure~\ref{fig:HalfTwinNetwork}.

While a regular deep neural network, denoted by LFT, for feature space reduction is not particularly novel, we emphasize the versatility of this approach, as both ontology and molecular feature for both drugs and proteins are reduced to similar dimensionality. This allows for a high amount of modularity and different experimental setups by plugging different kinds of features into the model. Additionally, these pretrained features can be used for a variety of other tasks. Additionally, the ontology LFT can be reused for a variety of DL2vec based features with respect to other ontologies and hypotheses. We hereby followed the results of \textit{DL2vec}, indicating that utilizing the activation function $\sigma := \mathrm{LeakyReLU}$ leads to performance increase.

@@

We included these molecular and ontology-based sub-models with a
larger graph convolutional model. The graph underlying the graph
convolutional neural network is based on the protein--protein
interaction (PPI) graph. The PPI dataset is represented by a graph
$G=(V,E)$, where each protein is represented by a vertex $v\in V$, and
each edge $e\in E\subseteq V\times V$ symbolizes an interaction
between two proteins. Additionally, we introduce a mapping
$x:V\rightarrow\mathbb{R}^{d}$ projecting each vertex $v$ to its node
feature $x_v := x(v)$, where $d$ denotes the dimensionality of the node
features.

% As described before, graph convolution has shown significant
% performance increase in a variety of tasks. While there are various
% methods out there we will only introduce the most basic one here. 
A graph convolutional layer \citet{GCNConv} consists of a learnable
weight matrix followed by an aggregation step, formalized by
\begin{equation}
	\mathbf{X}^{\prime} = \mathbf{\hat{D}}^{-1/2} \mathbf{\hat{A}}
	\mathbf{\hat{D}}^{-1/2} \mathbf{X} \mathbf{\Theta}
\end{equation}
where for a given graph $G=(V,E)$, $\hat{A} = A + I$ denotes the
adjacency matrix with added self-loops for each vertex, $D$ is
described by $\hat{D}_{ii} = \sum_{j=0} \hat{A}_{ij}$, a diagonal
matrix displaying the degree of each node, and $\Theta$ denotes the
learnable weight matrix. Added self-loops enforce that each node
representation is directly dependent on its own preceding one. The
number of graph convolutional layers stacked equals the radius of
relevant nodes for each vertex within the graph.

The update rule for each individual node is given by a message passing scheme formalized by
\begin{equation}
	\mathbf{x}^{\prime}_i = \mathbf{\Theta} \sum^{N}_{j}
	\frac{1}{\sqrt{\hat{d}_j \hat{d}_i}} \mathbf{x}_j
\end{equation}
where both $\hat{d}_i, \hat{d}_j$ are dependent on the edge weights
$e_{ij}$ of the graph. With simple, single-valued edge weights such as
$e_{ij}=1 \text{ }\forall (i,j)\in E$, all $\hat{d}_i$ reduce to
$d_i$, i.e., the degree of each vertex $i$. We denote this type of
graph convolutional neural layers with \textsc{GCNConv}.

While in this initial formulation the node-wise update step is defined
by the sum over all neighbouring node representations, we are able to
alter this formulation to another message passing scheme.  We are able
to rearrange the order of activation function $\sigma$, aggregation
$\mathrm{AGG}$ and linear neural layer $\mathrm{MLP}$ with this
formulation as proposed by \citet{GENConv2020}:
\begin{equation}
	\mathbf{x}_i^{\prime} = \mathrm{MLP} \left( \mathbf{x}_i +
	\mathrm{AGG} \left( \left\{
	\mathrm{\sigma} \left( \mathbf{x}_j + \mathbf{e_{ji}} \right) +\epsilon
	: j \in \mathcal{N}(i) \right\} \right)
	\right)
\end{equation}
where we will generally only consider
$\sigma \in \{\mathrm{ReLU}, \mathrm{LeakyReLU}\}$. We will denote
this generalized layer type as \textsc{GENConv}, following the
notation of PyTorch Geometric \cite{PytorchGeometric}.  While the
reordering is mainly import for numerical stability, this alteration
also addresses the vanishing gradient problem for deeper convolutional
networks \cite{GENConv2020}. Additionally, we can also generalize the
aggregation function to allow different weighting functions such as
learnable $\mathrm{SoftMax}$ or $\mathrm{Power}$ for the incoming
signals for each vertex, substituting the averaging step in
\textsc{GCNConv}. Hence, while \textsc{GCNConv} suffers from both
vanishing gradients and signal fading for large scale, highly
connected graphs, each propagation step in \textsc{GENConv} emphasizes
signals with values close to $0$ and $1$. The same convolutional
filter and weight matrix are applied to and learned for all nodes
simultaneously. % , and the resulting information\todo{Which information?
  % Specify} hold no information on their own connectivity.
We further employ another mechanism to avoid redundancy and fading
signals in stacked graph convolutional networks, using residual
connections and a normalization scheme \cite{DeepGCN2019,
  DeeperGCN2020}.
The residual blocks are reusable and can be stacked multiple times. %, thereby not losing focus of each node neighbourhood.
\todo[inline]{Needs a clearer outline/description/figure of YOUR model
structure.}
The structure is depicted in Figure~\ref{fig:ResGraphConvBlocks}.

@@

\subsubsection{title}
Combining half-twin and graph convolutional neural networks, we map all protein representations to the respective node features, initializing the graph convolutional update steps. The resulting representations are used for a similarity prediction similar to the models presented in figure ~\ref{fig:HalfTwinNetwork}.  
When combining ontology and molecular features with or without the
graph model, we concatenate both protein features and both drugs
features, before plugging them into the graph for the similarity
computation. A more in-depth image of the overall architecture, combining both feature types is depicted in figure ~\ref{fig:FullModelAllLayers}.

\subsubsection{Hyperparameter tuning}
As the number of drug-targets are sparse with respect to the number of
both drugs and proteins considered, the training, validation and
testing datasets are imbalanced. As there are only $22,336$ links in
the considered STITCH subset, the ratio
\begin{equation}
	w:= \frac{\#drugs \cdot \#proteins}{\#dti\_links} \approx 360,
	\label{weight}
\end{equation}
consequently needs suitable compensation in the computed loss function and
appropriate metrics for the evaluation.

Therefore, we weighted all positive drug-protein pair samples with
this ratio by introducing the following loss function with respect to
binary cross-entropy:
\begin{equation}
	l(x,y) = - w \left[ y \cdot \log x + (1 - y) \cdot \log (1 - x) \right]
\end{equation}
for a given prediction $x$ and target $y$, and positive weight $w$ defined by equation \eqref{weight}. We average this loss among
all drug-protein pairs in the training set, leading to a stable
environment for the used optimizing scheme \textit{Adam}
\citep{Adam2014}. We implemented a 5-fold cross validation split among the
proteins. Furthermore, we used early stopping in
the training process for optima detection.

To find the best hyperparameter configuration for the proposed model
we performed a grid search to find the most expressive and
non-redundant representation. We pretrained the bottom-up and the
top-down model separately and aimed at best performing models
with respect to our evaluation metrics. We optimized embedding sizes,
depth of the neural network, optimizer, learning rate and layer types
using an extensive, manual grid search, Starting from na\"ive, shallow feature transformations with an embedding size of 10, we scaled the network up to residual structures with up to 10 hidden layers leading to embeddings of size 4000, testing different network widths and learning rates for each configuration.

\subsection{Evaluation and metrics}

To assess each model, we compute a variety of common metrics for
binary classification. As the datasets are highly imbalanced, we use
the area under the receiver operating characteristic curve
(AUROC) on training, validation and testing split. % We
% calculated the true positive rate (TPR), false positive rate (FPR), and
% precision score. We will further compute true positives (TP), false
% positives (FP), false negatives (FN) and finally true negatives (TN).

We calculate the AUROC by computing true positive rate at various
false positive rate thresholds and using trapezoidal approximations to
estimate the area under the curve. We refer to this measure as
$MacroAUC$.

\todo[inline]{For RH: check this again later:}
In contrast we also calculate the $MicroAUC$ score. For given lists $D, P$ of drugs and proteins, respectively, and a set of known interactions $Int := \{(d_i, p_i) \}$, \textit{MicroAUC} is calculated as the average per entity (macro) \textit{AUROC} score. With respect to proteins, this can be formalized for given labels $l:D\times P \rightarrow \{0,1\}$ and predictions $y:D\times P \rightarrow [0,1]$  as 
\begin{equation*}
	MicroAUC_p := \underset{p\in P}{mean}\left(\left\{ \text{AUROC}(\{ (l(d_i, p), y(d_i,p))| d_i\in D\}) \right\}\right)
\end{equation*}

Drugs and proteins can be interchanged in this formulation, while we
refer to the different measures as protein-centric microAUC
($MicroAUC_p$) and a drug-centric microAUC
($MicroAUC_d$). Additionally, the $MicroAUC$ score may not be defined
as in some datasets some targets or drugs, respectively, do not have
any interactions, leading to an infeasible $TPR=0$ for all thresholds and an undefined
$AUROC$ score for that entity. For those entities we impute the $MicroAUC$ interpolating linearly, by using the accuracy for this subset.

@@

We developed \name as a computational model to predict drug--target
interactions. Specifically, given a protein, \name will identify and
rank drugs that likely target this protein. \name combines two types
of features: structural information for drugs and proteins that can be
used to determine if the drug and protein physically interact, and
information about phenotypic effects of drugs and changes in protein
function that may ``localize'' on an interaction network (i.e.,
neighboring nodes will share some of these features or be
phenotypically similar).  As structural features, \name uses
structural representations of drugs from the SMILES transformer
\cite{SmilesTransformer} and representations of protein amino acid
sequences from DeepGOPlus \cite{DeepGoPlus}.  \name learns
representations of drug effects and protein functions using the
ontology-based machine learning method DL2Vec \cite{DL2vec2020} and
ontology-based annotations of drugs and proteins.

We construct a PPI graph with proteins as nodes and protein-protein interactions as edges, mapping the protein features to each target as node features. \name then propagates information among the PPI network utilizing graph convolutional steps, calculating the similarity of drug and protein representation, and thus eventually predicting whether there is an interaction. The full workflow scheme is depicted in figure~\ref{fig:ModelWorkflow}

@@

\begin{table}[!tpb]
	\begin{tabular}{|p{2.2cm}|p{0.9cm}|p{0.75cm}|p{0.9cm}|p{0.75cm}|}
		\hline
		(a) STITCH results&\multicolumn{4}{c|}{PPI graph}\\
		&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
		&Macro AUC&Micro $AUC_p$&Macro AUC&Micro $AUC_p$\\
		\hline
		MolPred&$0.69$&$0.65$&$0.69$&$0.67$\\
		\hline
		OntoPred&$0.88$&$0.87$&$0.92$&$0.93$\\
		\hline
		\name (MolPred + OntoPred) & $0.89$ & $0.90$&$\mathbf{0.93}$&$\mathbf{0.94}$\\
		\hline
		\hline
		(b) Yamanishi results&\multicolumn{4}{c|}{PPI graph}\\
		&\multicolumn{2}{c|}{without}&\multicolumn{2}{c|}{with}\\
		&Macro AUC&Micro AUC&Macro AUC&Micro $AUC_p$\\
		\hline
		\name (MolPred + OntoPred) & $0.83$ & $0.82$&$0.84$&$0.84$\\
		\hline
	\end{tabular} \medskip

@@

	\caption{(a) + (b) Results for \name on STITCH and Yamanishi dataset evaluated with a 5-fold cross-validation. We hereby denote the molecular feature based predictor with \textit{MolPred}, while abbreviating the ontology based, top-down predictor with \textit{OntoPred}. (c) + (d) Results for various state of the art (c) drug--target interaction prediction methods on Yamanishi dataset and (d) drug--target affinity prediction methods on BIOSNAP dataset, evaluated on their original and the protein cross-validation splitting scheme, approximately reproducing the results of MolTrans \textbf{CITATION}.}
	\label{tab:Results}
\end{table}

@@

We trained, validated and finally tested all considered models on the
STITCH dataset using a 5-fold cross-validation over a protein
split; we then selected the best-performing models (with respect to
\textit{AUROC/MacroAUC} and $MicroAUC_p$) and evaluate them on in 5-fold protein split cross validation on the Yamanishi benchmark dataset to avoid validation overfitting and yield more realistic testing results. To evaluate the influence of the different
features separately, and determine whether they ``localize'' on the
PPI graph (and therefore can be exploited successfully by the graph
neural networks), we train and evaluate models with different types of
features, and with and without inclusion of the PPI graph,
separately, comparing the molecular (MolPred) and phenotype based (OntoPred) predictor and a combination of both, concatenating both features.
Table \ref{tab:Results} (a)+(b) shows the results of these experiments.
